---
title: "cms Vignette"
author: "Steven R. Talbot, Ines Koska, Heidrun Potschka"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{cms Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6
)
```

## About cms package
In our paper **"Design of composite measure schemes for comparative severity assessment in animal-based neuroscience research: a case study focussed on rat epilepsy models"** we introduce composite measures schemes (cms) to assess severity in rat epilepsy models. We suggest a PCA-based algorithm to assess a plethora of outcome variables for their suitability as features in a simple composite score for comparative severity assessment.

This vignette covers the basic functions for cms analysis in R [@RCoreTeam].

# Package Requirements
In order to use the cms package some dependencies need to be installed. The ade4 and made4 packages are an essential requirement for the cms 
algorithm. However, they are available from Bioconductor. They can be installed as such:

[BiocManager](https://cran.r-project.org/web/packages/BiocManager/vignettes/BiocManager.html)

```
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("made4")
```
The remaining dependencies can be loaded via CRAN: ade4 [@ade4_1, @ade4_2, @ade4_3, @ade4_4], scatterplot3d [@sp3d], 
lattice [@lattice], corrplot [@corrplot2017], caret [@caret], ggplot2 [@ggplot2], reshape2 [@reshape2], plyr [@plyr],
RColorBrewer [@colbrew], gplots [@gplot], e1071 [@e1071]. The following links can be used directly in the R console.

```
install.packages("ade4") 
install.packages("scatterplot3d")
install.packages("lattice")
install.packages("corrplot")
install.packages("caret")
install.packages("ggplot2")
install.packages("reshape2")
install.packages("plyr")
install.packages("RColorBrewer")
install.packages("gplots")
install.packages("e1071")
```
Please note that the cms package requires graphical components which are not fully used in the current version of the development version (v. 0.0.9000). 

See the [reference section](#references) for more information on the included packages. 

# Code and information
For more information on the cms package and the source code see:  

* cms [website](https://talbotsr.com/cms/index.html) and  
* the cms [GitHub repository](https://github.com/mytalbot/cms)  

# Introduction 
According to the EU directive 2010/63 it is essential to assess the severity of interventions and models in experimental animals.
Therefore, sensitive and robust parameters are crucial. Based on the data set generated in three different epilepsy models promising
parameters were selected and combined to a composite measure scheme. The cms package guides through the process of selecting potential
parameters or to allocate your animals to clusters based on the cms. Related to the predefined or individual selection and combination of
parameters the respective clusters and the distribution of the animals to these clusters will be visualized.  

The calculated clusters describe different levels of severity. Please consider that the best parameters can vary for different models,
species or even for different laboratories. Therefore, this tool can help to select the most interesting parameters from your individual,
generated data.  

For the interpretation of the different severity levels you have to consider that they are based on the highest severity animals
experienced in your experiments but not necessarily the highest existing. Moreover, there might always be “better” parameters. So be aware
of the limitations of your individual created cms. In conclusion, the cms package is a useful tool to get a retrospective overview of the
severity your animals experienced during the experiment. Therefore, it can help to complement a mostly empiric severity assessment. 

**What the cms package...**

|  ... **can** do | ... **can’t** do |
|:-----------------------------------|:-----------------------------------|
| Helps to select interesting parameters for severity assessment | Interpretation of the selected parameters in terms of how useful they can be to assess the impairment of animals’ wellbeing |
| Provides a basis for quantitative and comparative severity assessment (incl. comparison with data from the different epilepsy models) | |
| Allocation of individual animals to a severity level | Direct translation of the different severity levels to those of the EU directive 2010/63 | 
| Be a useful tool to complement an empiric severity assessment in an objective and scientific way | Replace the common severity assessment |

# Data structure
The cms algorithm requires the input data in specific long format. The table can be combined with meta information. The user has to specify a column from which on only measurement data are included. Before that column all entries can have informative character.

Meta-information (=column headers) in the included data set are:  

* **shortname**   a short term study identifier (e.g., kindling)
* **for2591_id**  internal identifier  
* **species**     analysed species
* **sex**         analysed sex
* **strain**      animal strain id
* **animal_id**   animal identifier (mandatory!)
* **treatment**   experimental treatment (mandatory!)
* **subproject**  subproject (mandatory if there are subprojects!)
* **groups**      subgroups (mandatory if there are additional subgroups in the data)
* **mod**         model string (e.g., chem, elec, kind)
* **repeats**     experimental repeats (mandatory!); if only one repeat set this column to 3

Please note that most of the presented meta-data are optional. However, you may use any of the included criteria to filter the data
for subsets.

Measurement data in the provided example start at column No. 14. The following 57 variables are included. Note that the raw data include a lot of missing values (NA). These will be curated in the algorithm.

Bwc, rgs, tint_mouse, nesting_rat, burrowing_rat,  openfield_rat, cort_plasma, cort_fec, cort_fur_hair, food, social_interaction,
Sacc_pref, OF_rearing_frequency, Of_immobile_duration,  OF_center_time, BWB_WB, BWB_entries_in_BB, BWB_streching, BWB_LT, EPM_stretching,
EPM_head_dip, EPM_closed_arms_duration, EPM_open_arms_duration, EPM_open1_3, Adrenal_glands, Creatincinase, Oxytocin, BDNF, Irwin,
Seizures_n, Seizures_duration, Act_d, Act_n, HR_d, HR_n, NN.I_d, NN.I_n, NNx_d, NNx_n, pNNx_d, pNNx_n, RMSSD_d, RMSSD_n, SDNN_d, SDNN_n,
MPPF_hippocampus_L, MPPF_hippocampus_R, MPPF_mPFC, MPPF_septum, FDG_mPFC, FDG_septum, FDG_amygdala, FDG_striatum, FDG_hypothalamus,
FDG_thalamus, FDG_parietal_cortex, FDG_hippocampus 

The full data set is included in the cms package and can be accessed with:

```
library(cms)
head(episet_full) 
```

# Using the functions of the cms package
The following section explains the functions of this package and how to use them.

## cms_load
Data from a *.txt `file` will be read into R using the `cms_load` function so that it can be fed to the algorithm. The data should comprise of meta-information for sub-sampling and a general experimental overview. After the meta-section the variables should follow. The user has to specify the column number later that points to the first variable after the meta-section. The cms_load function requires a "repeats" column that is set to 3. If you do not have repeats, just add a column and fill its contents with 3. Sorry, we will address this little imperfection in a later step...
The `remove_this` object can be used to remove specific columns from the data table (e.g., `remove_this=c("Init.SD", "total.SD", "Mean.SD")`). This is good for additional cleaning of the data set.

```
# This function loads tabular text data into R.

cms_load(file, remove_this=NA)

```

## cms_clusters
The algorithm takes the cleaned input data (`raw`), does some curation for missing values (`emptysize=0.2`, e.g., for eliminating data with more than 20 percent missing entries) and then calculates the composite score. The feature selection is done be re-sampling random sets of the data and monitoring the outcome of the respective PCA analysis. The fractional size of the sampling set is called `trainsize`. The number of sampling repeats is coded in `runs`. The algorithm requires individual animal ids. The column header can be coded differently. Here, it is called `idvariable="animal_id"`. The first variable for the analysis after the meta-information starts at colum 14 and is coded in `varstart`. The algorithm will analyse the full data table, unless a limit is set in `varend`. Again, there is an option to selectively exclude columns from the analysis (`exclude`).

Additionally, the function can be used in two ways:

1. with the object `scorevars` set to NA the data are screened for the most prominent feature contributions
2. the so found feature sets can be specified, e.g., `scorevars=c("Sacc_pref", "social_interaction", "burrowing_rat", "openfield_rat")`.
This will result in the calculation of a severity-based cluster distribution analysis using the input variables.


```
# This function calculates the cluster and frequency distribution of composite measures for severity attribution. 

cms_clusters(raw= raw, runs=100, emptysize=0.2, trainsize=0.8, idvariable="animal_id", 
varstart=14, varend=NA, exclude="Seizures_n", scorevars=NA)

```

## cms_analysis
The function takes in the results from the cms_clusters function (frequency distributions of each run of the 80 percent subsampling of the
training data) and calculates general attributes as well as graphical distribution outputs. Again, the `raw` data are required as inputs. The `thresholds` variable is the first output from the `cms_clusters` function and is itself a list. The object `cluster_distribution` is the second output of the `cms_clusters` list and includes the frequency distributions of the most prominent features. If `savepath=NA`, no plot will be saved. Otherwise, if a path is given a cluster distribution plot for severity analysis of the subgroups will be stored at the location. 

```
# This function analyses the output of the specific cms_cluster function.

cms_analysis(raw, thresholds, cluster_distribution, savepath=NA)

```
# Example Analyses 

### Example 1: Identification of the most prominent (stable) features in the provided data set `episet_full`.

```{r setup}
suppressPackageStartupMessages(library(cms))

cms_cl        <- cms_clusters(episet_full, 
                              runs        = 100, 
                              emptysize   = 0.2, 
                              trainsize   = 0.8, 
                              idvariable  = "animal_id", 
                              varstart    = 14, 
                              exclude     = "Seizures_n",
                              scorevars   =  NA) 
cms_cl
```


### Example 2: Specific feature analysis for severity comparisons.
In the resulting plots, darker colors refer to higher severity.

```{r setup 2}
suppressPackageStartupMessages(library(cms))

cms_cl        <- cms_clusters(episet_full, 
                              runs        = 100, 
                              emptysize   = 0.2, 
                              trainsize   = 0.8, 
                              idvariable  = "animal_id", 
                              varstart    = 14, 
                              exclude     = "Seizures_n",
                              scorevars   =  c("Sacc_pref", "social_interaction", "burrowing_rat", "openfield_rat")) 

thresholds           <- cms_cl[[1]] 
cluster_distribution <- cms_cl[[2]]  

result               <- cms_analysis(episet_full, thresholds, cluster_distribution)
result[[2]]
```




# References




