---
title: "cms Vignette"
author: "Steven R. Talbot, Heidrun Potschka"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{cms Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(include = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(cms)
library(knitr)
library(dplyr)
library(kableExtra)
```


## About cms package
In our paper"Design of composite measure schemes for comparative severity 
assessment in animal-basedneuroscience research: a case study focussed on rat 
epilepsy models" [@Dijk2020], we introduce composite measures schemes (cms) to
assess severity in rat epilepsy models. We suggest a PCA-based algorithm to
assess a plethora of outcome variables for their suitability as features in a 
simple composite score for comparative severity assessment.

This vignette covers the basic functions for cms analysis in R [@RCoreTeam].

# Package Requirements
In order to use the cms package, some dependencies need to be installed. 
Please make sure to install the following R-packges from CRAN before using cms.

ggplot2 [@ggplot2]  
dplyr [@dplyr]
factoextra [@factoextra]

```
install.packages("ggplot2") 
install.packages("dplyr")
install.packages("factoextra")
```
# Code and information
For more information on the cms package and the source code see:  

* cms [website](https://talbotsr.com/cms/index.html) and  
* the cms [GitHub repository](https://github.com/mytalbot/cms)  

# Introduction 
According to the EU directive 2010/63 it is essential to assess the severity of 
interventions and models in experimental animals. Therefore, sensitive and
robust parameters are crucial. Based on the data generated in three 
different epilepsy models, promising parameters were selected and combined to a
composite measure scheme. The cms package guides through the process of
selecting potential parameters or to allocate your animals to clusters based on 
the cms. Related to the pre-defined or individual selection and combination of 
parameters, the respective clusters and the distribution of the animals to these 
clusters will be visualized.  

The calculated clusters describe different levels of severity. Please consider 
that the "best" parameters can vary in different models, species or even between
different laboratories. Therefore, this tool can help selecting the most 
interesting parameters from individually generated data.  

For the interpretation of the different severity levels, one has to consider
that they are based on the highest severity animals experienced in the 
experiments but not necessarily the highest existing. Moreover, there might
always be “better” parameters. So be aware of the limitations of the 
individually created cms.

In conclusion, the cms package is a useful tool to get a retrospective overview
of the severity in animal data experienced during an xperiment. Therefore, it
can help to complement a mostly empiric severity assessment. 

**What the cms package...**

|  ... **can** do | ... **can’t** do |
|:-----------------------------------|:-----------------------------------|
| Helps to select interesting parameters for severity assessment | Interpretation of the selected parameters in terms of how useful they can be to assess the impairment of animals’ wellbeing |
| Provides a basis for quantitative and comparative severity assessment (incl. comparison with data from the different epilepsy models) | |
| Allocation of individual animals to a severity level | Direct translation of the different severity levels to those of the EU directive 2010/63 | 
| Be a useful tool to complement an empiric severity assessment in an objective and scientific way | Replace the common severity assessment |

# Data structure
The cms algorithm requires the input data in specific long format. The table can be combined with meta information. The user has to specify a column from which on only measurement data are included. Before that column all entries can have informative character.

Meta-information (=column headers) in the included data are:  

* **shortname**   a short term study identifier (e.g., kindling)
* **for2591_id**  internal identifier  
* **species**     analysed species
* **sex**         analysed sex
* **strain**      animal strain id
* **animal_id**   animal identifier (mandatory!)
* **treatment**   experimental treatment (mandatory!)
* **subproject**  subproject (mandatory if there are subprojects!)
* **groups**      subgroups (mandatory if there are additional subgroups in the data)
* **mod**         model string (e.g., chem, elec, kind)
* **repeats**     experimental repeats (mandatory!); if only one repeat set this column to 3

Please note that most of the presented meta-data are optional. However, you may
use any of the included criteria to filter the datafor subsets.

Measurement data in the provided example start at column No. 14. The following 
57 variables are included. Note that the raw data include a lot of missing 
values (NA). These can be curated in the algorithm using the `emptysize` object.

Bwc, rgs, tint_mouse, nesting_rat, burrowing_rat,  openfield_rat, cort_plasma, 
cort_fec, cort_fur_hair, food, social_interaction, Sacc_pref, 
OF_rearing_frequency, Of_immobile_duration,  OF_center_time, BWB_WB,
BWB_entries_in_BB, BWB_streching, BWB_LT, EPM_stretching, EPM_head_dip, 
EPM_closed_arms_duration, EPM_open_arms_duration, EPM_open1_3, Adrenal_glands, 
Creatincinase, Oxytocin, BDNF, Irwin, Seizures_n, Seizures_duration, Act_d, 
Act_n, HR_d, HR_n, NN.I_d, NN.I_n, NNx_d, NNx_n, pNNx_d, pNNx_n, RMSSD_d, 
RMSSD_n, SDNN_d, SDNN_n, MPPF_hippocampus_L, MPPF_hippocampus_R, MPPF_mPFC, 
MPPF_septum, FDG_mPFC, FDG_septum, FDG_amygdala, FDG_striatum, FDG_hypothalamus,
FDG_thalamus, FDG_parietal_cortex, FDG_hippocampus 

The full data are included in the cms package and can be accessed with:

```
library(cms)
head(episet_full) 
```

# Minimum data 
The cms function requires a minimum on information to work. Depending on the 
number of subsets, the mandatory table fields are:  

animal_id  
treatment (e.g., "naive", "sham" and "SE")  
mod (e.g., "chem", "elec" and "kind")  
variables (e.g., burrowing, corticosterone, openfield,...)  

The treatment and mod columns can be used for subsetting or pooling the data.

# Example 1
Let's analyze repeat No. 3 from the episet_full and use the a battery of 23
variables in the data to extract features. Note, that this analysis uses
the pooled treatment ("naive", "SE" and "sham") and mod ("chem", "elec" and 
"kind") argument. 


```{r echo=TRUE}
# first we select only the repeats = 3
set       <- cms::episet_full[episet_full$repeats %in% 3, ]

# Now the variables that shall be testes are called in a vector
variables <- c("Sacc_pref", "social_interaction", 
               "burrowing_rat", "openfield_rat")

# do the cms feature analysis
features  <- cms(raw        = set,
                 runs       = 300,
                 idvariable = "animal_id",
                 setsize    = 0.8,
                 variables  = variables,
                 maxPC      = 4,
                 clusters   = 3, 
                 showplot   = FALSE,
                 legendpos  = "none")
```


```{r echo=TRUE, fig.cap="Feature distributions after cms analysis."}
features$p
```


The plot is rather crowded so that the information we are seeking is better 
extracted from the attached frequency (features$FRQ) table. Only looking at the
first 10 rows.

```{r echo=FALSE}
features$FRQ %>%
  slice(1:10) %>%
  kbl() %>%
  kable_styling() 

```

The result shows the following feature order in repeats=3, with the above 
listed variables, using 4 principle components in 100 repeated runs:
  BWB_WB (40\%) > BWB_LT (32\%) > Adrenal_glands (20\%) > cort_fec (6\%) > 
  openfield_rat (1\%) > social_interaction	(1\%).

The remaining variables were not selected and/or not contributing to the PCA. 
Therefore, they were not deemed to be relevant features.

# Cluster analysis
The cms function also calculates a composite score from the principal component
scores. This "compscore" is simply the sum of the 1 to `maxPC` PCA scores. The
result is stored in the result of the cms function as a list object. In the 
upper example, we can access this information with 
features$reportdata$compscore. 

The observant reader might have noticed the clusters=3 object in the cms
function. This is the number of clusters that shall be applied to the 
"compscore". Unfortunately, this is a heuristic. This means: the user will have
to figure out how many clusters are correct for the data. The cluster 
information, or better: the distribution of the clusters are characteristic for
specific subgroups. 

Fortunately, the cms function takes care of the heuristic determination of the
number of clusters. The weighted sums of squares are calculated with the 1 to 
`maxPC` PCA scores of the data and the selected number of clusters is marked 
with a vertical line. With the Kaiser-Guttman criterion ("ellbow method") the
ideal number of clusters can be inferred. 

```{r echo=TRUE, fig.height = 3, fig.cap="With the scree plot the number of clusters can be determined."}
features$pscree
```

Now it becomes clear, why we have kept the subgroup information in the feature
extraction above. We can easily control the subgroups with the dplyr package and
plot the cluster distributions. The plot_cms function will take care of this.

```{r echo=TRUE, fig.height = 5, fig.cap="Cluster frequencies after cms analysis."}
cms_clustering <- plot_cms(cmsresult = features, 
                           rotateX   = 45)
cms_clustering$p
```


```{r echo=FALSE}
cms_clustering$clustering %>%
  kbl() %>%
  kable_styling() 

```


# References




